Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-05-03 Paper/paper.tex

\Ken{[Guys---I recommend we get a better name than confinement.  It's totally uninformative.  In my view, better would
    be `confine-and-configure' or something.  Please give it some thought.]}

\Justin{[Ken---I agree that the name isn't great, but I don't think we should change it because: it's already in the
    literature and we didn't come up with the method. I think we should consider capitalizing it as Confinement Method
    throughout.]}

\section{Introduction}

In some problems of protein science, you want to know the relative stability of a protein's conformation $A$ compared to
its conformation $B$.  We call this the \emph{difference free energy}. For example, in all esoteric mechanisms, a protein
often adopts one conformation when a ligand is bound and another conformation when no ligand is bound~\cite{Elber2007}.
To understand the mechanism requires knowledge of the relative free energies of the protein conformations.  Another
example is in protein structure prediction, for example as practiced in the Critical Assessment of Structure Prediction
(CASP) competition~\cite{Moult2011}. If you are using some computational model and have predicted two putative native
structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which is the more
native-like.  In a third example, you may want to know whether mutating a few amino acids in a protein could cause the
protein to switch from one stable conformation to another, because that could have important consequences for biological
mechanisms and disease. And fourth, sometimes binding a ligand induces a protein from one conformation to another. A
quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free energy of the
protein in the two states. In all these cases, it would be useful to have a computational method that can be used
efficiently with atomically detailed physical forcefield models to compute the free energy differences between different
given protein conformations.

\Justin{[TI is actually the basis of the confinement step. We need to be more explicit about what we mean here.] Methods
    like thermodynamic integration~\cite{Tironi1994},\cite{Meirovitch2007} which are successful in alchemical free
    energy calculation may not be as useful for calculation of conformational free energy.} One widely explored strategy
is to use molecular dynamics simulations along some putative reaction coordinate pathway from conformation $A$ to
$B$~\cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006}. \Ken{[we need many refs here: cite
    Elber's milestoning, Jhih-Wei Chu, and the many other path-sampling approaches.]} The free energy along this
reaction coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977}\Justin{[add ref
    where this is used computing free energy differences between conformations]} or targeted MD \Justin{[refs]}. Such
approaches have several limitations. First, it is necessary to know an efficient reaction pathway from $A$ to $B$. If
conformations $A$ and $B$ are quite different, then it can be challenging to find such paths. Second, these methods are
computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires
accurate determinations of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$,
so each step requires substantial amounts of sampling.  And third, these methods are prone to large errors, because the
pathway error is a sum of many errors among the many steps.

\Ken{[Are all these methods below ones that don't involve paths?  Also, are there other key methods we should cite?  We
    need a good scholarly list of all the relevant works.]}

The calculation of protein conformational free energy has been successfully attempted by a number of groups
\cite{Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}. Some of
these methods like the reference system method~\cite{Ytreberg2006}, deactivated morphing~\cite{Park2008}, the
confinement method~\cite{Tyka2006,Cecchini2009,Ovchinnikov2013} take an alternative strategy for computing difference
free energies that does not require knowing a pathway from $A$ to $B$.  \Ken{[please add the several new Karplus papers
    from 2012-2013.]}.

Here, we adopt the confinement method of Tyka et al~\cite{Tyka2006} and Cecchini et al.~\cite{Cecchini2009}, which is
based on the thermodynamic cycle shown in Figure~\ref{fig:method}. We start with an ensemble of related conformations,
$A$. We then restrict $A$ to a much ``tighter'' ensemble $A^\ast$ by applying position restraints in a series of MD
simulations. Similarly, a second ensemble $B$ is restricted to $B^\ast$. We then compute the free energy between
$A^\ast$ and $B^\ast$ using either normal mode analysis or the quasiharmonic method [ref]. This step takes into account
both the remaining conformational entropy in each ensemble and the remaining enthalpy difference. In this way, we can
calculate the free energy difference between the two end states without needing to define a physical path or reaction
coordinate connecting the two states. The confinement method shares some similarities with the `confine-and-release'
method for computing ligand binding affinities~\cite{Mobley2007,Mobley2012,Mobley2006} \Ken{[cite mobley's jctc 07 (ref
    216 on our labsite), and mobley's recent j chem phys review, and mobley (ref 199) j chem phys, and others that are
    relevant.]}.

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{method.pdf}\label{fig:method}
    \end{center}
    \caption{Graphical representation of the thermodynamic cycle involving confinement method.}
\end{figure}

Previously, the confinement method has been validated on small model peptides~\cite{Tyka2006,Cecchini2009}. Here, we
show that the confinement method can also be applied successfully to larger systems, such as the series of chameleon
proteins designed by Orban and co-workers~\cite{Alexander2007,He2008,Alexander2009,He2012}, which can switch between two
completely different folds with only slight changes in sequence. We also show that our computed difference free energies
are useful in evaluating the quality of CASP target protein predictions.  This could be of significant value for
ultimately improving the energetics in protein-structure models, not just the structures. And, finally, we show that we
can approximately decompose difference free energies into individual amino acid components. This offers the opportunity
for diagnosing the structural basis for difference free energies, which can be useful for interpreting biological
mechanisms.

\section{Results and Discussion}

\subsection{The confinement method succeeds at some basic consistency checks}

@@

Second, we looked at 6 target proteins from the CASP9 experiment [ref].  For each target, we examined up to 5 submitted
models.  We computed the difference free energy between the true native and the best model.
Figure~\ref{fig:summary_casp} and Table~\ref{table:casp_control} show that, in 5 out of 6 cases, the confinement method
assigns a lower free energy to the experimentally determined structure than to any of the decoys.  Other well-known
discriminators can also successfully tell native structures from computer generated models~\cite{Sheffler2009} [add
rosetta holes, rosetta ff, dfire]; it is just an independent useful validation that the confinement calculations make
sense.

\Justin{[Arijit, please add GDT of best decoy to this table.]}

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally determined structure than to
    computer-generated predictions. For each target, we examined as many as five predictions submitted by CASP
    participants. We report the free energy difference between the most favorable decoy and the experimentally
    determined structure. Positive $\Delta\Delta G$ values indicate that the experimental structure is predicted to be
    more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G = G_{best~decoy} - G_{native}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}
\end{table}

\subsection{The confinement method correctly predicts the structures of chameleon sequences}

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.!\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}. These are instances in which two highly
similar sequences fold into remarkably different structures. Orban and co-workers have designed a protein-G-like
sequence of 56-residues that is marginally stable in one of two possible folds. By mutating key residues in this
sequence they are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta +
\alpha$ structure as the $\beta$ conformation, and we refer to the $3\alpha$ structure as the $\alpha$ conformation. One
pair of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair
(GA95/GB95) is 95 percent identical, differing in three positions. Accurately predicting the structural preferences of
these structures presents a serious challenge for computational
methods~\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}. \Justin{Do any of these citations try to
    do any modeling or comment about how challenging this would be?}

\begin{figure}
    \begin{center}
    \includegraphics[width=5.0in]{orban.pdf}
        \label{fig:orban}
    \end{center}
    \caption{The confinement method correctly predicts the structural preferences of four chameleon sequences. The top
        part of the figure represent four sequences used in this study along with the protein data bank identifier. The
        experimentally observed fold, computationally predicted fold, and difference free energy between the two folds
        is reported for each sequence.}
\end{figure}

@@

\subsection{Per-residue free energy calculations can identify the mechanistic detail behind conformational preferences}
\label{sec:orban_per_residue}

To better understand the mechanism behind these chameleon proteins, we decomposed the calculated free energy into
per-residue contributions in an approximate way~\cite{Tyka2006}. We decompose each confinement step ($\Delta
G_{A,A^\ast}$, $\Delta G_{B,B^\ast}$) into per-residue contributions. We can also decompose the remaining enthalpy in
the confined ensemble into per-residue contributions. However, the method is approximate because we do not include the
residual conformational entropy from the normal mode or quasiharmonic analysis steps. Computing per-residue
contributions helps us to identify important residues that stabilize a particular conformation. The per residue free
energy, $\Delta \Delta G ((\beta) - (\alpha))$ is shown in Figure~\ref{fig:perresidue_orban}.

\begin{figure}
    \begin{center}
        \includegraphics[width=4.5in]{delta_g.pdf}
        \label{fig:perresidue_orban}
    \end{center}
    \caption{Per-residue difference free energies for the $\alpha$ and $\beta$ conformations of the GA95 and GB95
        sequences. Residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure,
        and white residues have no preference.}
\end{figure}

Although the overall free energy difference between the two structures is small (within 3--5 kcal/mol), individual
residues can show marked preferences for being in either $\alpha$ or $\beta$ conformation. Such differences can be
understood by looking in detail at the local environment for those residues. For example, the region around residue 7
forms a random coil in the $\alpha$ structure, whereas it is forming a beta sheet structure in the $\beta$ structure.
These residues strongly favour the locally well packed and hydrogen bonded environment found in the $\beta$ sheet.
Overall, favoring $\alpha$ or $\beta$ structure is a delicate balance, were the relative global free energy difference
is small and the contributions of different per residue tendencies balance out. It is therefore very likely that by
changing key residue preferences the global fold preferences can be changed.

@@

\subsubsection{The confinement method can correctly rank-order structure predictions}

First, we examined the ability of the confinement method to rank different models for a target that have been generated
by a single group using the same methodology for all predictions. We examined two targets: T0559 and T0560 (see Table
?).

\Justin{[Arijit: Let's move the protein names, organisms, pdb ids, etc to a separate table and just use the CASP ids.]}

The first test case is target T0559. The best predictor group for this 69 amino acid target was ``BAKER-ROSETTASERVER''.
To save computer time, we excluded two models that were very similar to other models that we did include. For this
target, the difference free energy computed by the confinement method can be used to accurately rank-order all of the
models and the native structure (Figure~\ref{fig:T0559}).

\Ken{Arijit- we could probably make a better visualization of this figure above by putting it onto a toy landscape:
    x-axis is rmsd, y-axis is DDG, and above each of the 4 points on this graph will be the ribbon diagram and the
    GDT-TS.  I suggest we do that with the fig below too.  Let's don't use 4 different colors for the fig below.  Let's
    just show differences in structures in some different color, or show them all as green.]}

\Justin{Ken: I don't mind the 3d plots that are in here now, but I'm not sure I would like something that looks like a
    free energy landscape.}

\begin{figure}
    \begin{center}
        \includegraphics[width=3.8 in]{T0559.pdf}
        \label{fig:T0559}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and three predictions produced by a single
        method.}
\end{figure}

\Justin{I'm not sure we need a second figure of this. Move it to the SI?}

We performed a similar calculation for target T0560 with two models form the group ``Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
sate and our calculated free energy based ranking matches well with GDT-TS (Figures S?).

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{T0560.pdf}
    \end{center}
    \caption{Native structure and two models (from group ``Splicer'') for target T0560.}
    \label{fig:T0560}
\end{figure}

@@

\subsubsection{The per-residue free energy is sensitive to small changes in protein conformation}

In section~\protect\ref{sec:orban_per_residue} we discussed how the per-residue free energy can reveal mechanistic
detail behind the conformational preference of a chameleon sequence. In this section, our aim is to apply the same
method and try to understand if the per-residue free energy can help us identify residues which stabilize or destabilize
a particular region of a protein. In this case, the two conformations have similar folds with only small changes in
localized regions. We chose CASP target T0569 and compared the experimental NMR structure with the best predicted model
(GDT-TS=78; predicted by the ``Mufold'' group). The confinement method predicts that the native structure is more stable
by 20 kcal/mol (Figure S2).

The confinement method clearly identifies two hydrophobic residues V59 and I61, which destabilize the predicted model
with respect to the crystallographic structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The sidechains of
these hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but oriented
towards the exterior of the protein and are solvent exposed in the model. These residues also form part of a beta-sheet
in the experimental structure, but do not form inter-strand hydrogen bonds in the predicted model. There is also a large
difference around K76, which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure.
This suggests that salt-bridge interactions are too favorable for the combination of force field and implicit solvent
model we use, which has been a problem noted in the past [ref carlos].

\Justin{[Arijit: Figure~\ref{fig:T0569_per_residue} says crystal, but the text says NMR.]}

@@

\subsubsection{The confinement method occasionally produces incorrect results}

\Justin{Arijit: let's move Figure~\ref{fig:T0538} to the supporting info and move Figure S4 into the main text.}

Despite success for most of the studied systems, there are few failures, specially when the GDT-TS scores of the
compared structures are very close. One example is Target T0538, where we compared the crystal structure with three
models (Model 1: ``PconsR''---GDT-ts=96; Model 2: ``Shell''---GDT-TS=90; Model 3: ``FOLDIT''---GDT-TS=86). Contrary to
our expectation, the confinement method predicts that Model 1 is more stable than the crystal structure (Figure ?).
Per-residue free energy calculations (not shown) show that despite only small variations at the backbone level, the
sidechains are oriented in very different ways (Figure ?), giving rise to large differences in the stabilization of
certain residues. In particular, some of the differences arise from different salt bridge patterns and certain flexible
polar residues exposed to the surface. This unexpected result shows that the confinement method is very sensitive to
local interactions (including sidechain reorientation) and may indicate issues with the forcefield and implicit solvent
models used in the calculation.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{T0538.pdf}
\end{center}
\caption{The native and three model structure of engineered protein from Asr4154 protein (PDB ID: 2L09 and CASP
    code:T0538). The model 1,2 and 3 are from the group PconsR, Shell and FOLDIT respectively.}
\label{fig:T0538}
\end{figure}

\subsubsection{What can we say about low resolution models?}

\Justin{I think we should consider dropping this whole section. We have a sample size of one and I don't expect that we
    can generally do well ranking models with a GDT of 44. I think this is a fluke.}

So far we have seen that the method is good at predicting preferences when the structures are not very far from the
native. But the question remains how far from native can we go and still see that the method produces correct result. In
this section we explored this question with models from  extracellular domain of the jumping translocation breakpoint
protein (pdb id: 2KJX). Most of the group could only generate low resolution models for this CASP9 target (id: T0531) .
In our comparison, we choose five models by the group MUFOLD-MD, which was the best performing group for this target
with their best model had a GDT\_TS value of 44. The result presented in Figure~\ref{fig:T0531} shows: 1.) native is
correctly identified as expected and 2.) Surprisingly there is a high level of correlation between the GDT ranking and
the free energy ranking for model 1 and model 3, the rest three structures with GDT\_TS score less than 35  are ordered
incorrectly. It is worth to note that models 2 and 3 have the same GDT and very different free energies, meaning that
the actual ordering could change a lot \cite{Perez2012}. It is encouraging that at least the method can pick out the
best model even though it is got a low GDT score: 44.

\begin{figure}
\begin{center}
\includegraphics[width=3.5 in]{T0531.pdf}
\end{center}
\caption{The native structure and 5 models of extracellular domain of the jumping translocation breakpoint protein (pdb
    id: 2KJX and the CASP code: T0531).}
\label{fig:T0531}
\end{figure}

\subsubsection{Can the confinement method perform better quality assessment in protein structure prediction?}

\Justin{Arijit---is this the case where we chose a structure that was missed by the consensus predictors, if so, we
    should try to indicate this. The point of this section should be that although consensus prediction strategies do
    well, they can miss a good answer that is only predicted by a few methods. At least in this case, the confinement
    method is able to capture this structure that was otherwise missed.}

A part of the CASP experiment is dedicated to the quality assessment (QA) of predicted models \cite{Kryshtafovych2011}.
Here, predictors were asked to score each model on a scale (known as qmode) from 0 to 1, with higher values
corresponding to better models \cite{Kryshtafovych2011}. It will be interesting to know how confinement method perform
in quality assessment compare to the other groups in CASP9. We investigated this using couple of CASP Targets. Here, we
present a case, where confinement method perform well than the top performing group MUFOLD-WQA \cite{Wang2011} of CASP9.
We choose  two models from CASP taget T0538. They were model 3 submitted by PconsR (GDT\_TS = 96, qmode 1 = 0.5434) and
model 5 from the MULTICOM-NOVEL ( GDT\_TS = 83, qmode 1 = 0.5865). Both are server predicted models and the qmode value
presented are from MUFOLD-WQA \cite{Wang2011}. We choose these two models as the model by PconsR was the most accurate
predicted model for this target, whereas the other model was predicted best by QA test of MUFOLD-WQA. The calculated
free energy using confinement method indicate that the model by PconsR is more stable by $3.9 Kcal/Mol$ which support
the GDT\_TS trend. There is no doubt that the consensus approach can predict the model quality in a faster manner. But
we expect confinement method can predict the model quality in a relatively expensive but much more accurate way.

\section{Conclusion}

@@

\item  In the first step, a minimization of A and B conformations are performed. These minimized conformations
    ($A^{\ast\ast}$ and $B^{\ast\ast}$) are the reference conformation of that state.

>>>>

\section*{Introduction}

In some problems of protein science, you want to know the relative stability of a protein's conformation $A$ compared to
its conformation $B$.  We call this the \emph{difference free energy}. For example, in all allosteric mechanisms, a protein
often adopts one conformation when a ligand is bound and another conformation when no ligand is bound~\cite{Elber2007}.
To understand the mechanism requires knowledge of the relative free energies of the protein conformations.  Another
example is in protein structure prediction, for example as practiced in the Critical Assessment of Structure Prediction
(CASP) competition~\cite{Moult2011}. If you are using some computational model and have predicted two putative native
structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which is the more
native-like.  In a third example, you may want to know whether mutating a few amino acids in a protein could cause the
protein to switch from one stable conformation to another, because that could have important consequences for biological
mechanisms and disease. And fourth, sometimes binding a ligand induces a protein from one conformation to another. A
quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free energy of the
protein in the two states. In all these cases, it would be useful to have a computational method that can be used
efficiently with atomically detailed physical forcefield models to compute the free energy differences between different
given protein conformations.

One widely explored strategy
is to use molecular dynamics simulations along some putative reaction coordinate pathway from conformation $A$ to
$B$~\cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006,Elber2005,Chipot2007}. The free energy along this
reaction coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977,Mascarenhas2013}  
and weighted histogram analysis method (WHAM) \cite{Kumar1992}. Such
approaches have several limitations. First, it is necessary to know an efficient reaction pathway from $A$ to $B$. If
conformations $A$ and $B$ are quite different, then it can be challenging to find such paths. Second, these methods are
computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires
accurate determinations of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$,
so each step requires substantial amounts of sampling.  And third, these methods are prone to large errors, because the
pathway error is a sum of many errors among the many steps.

The calculation of protein conformational free energy has been successfully attempted by a number of groups
\cite{Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}. Some of
these methods like the reference system method~\cite{Ytreberg2006}, deactivated morphing~\cite{Park2008}, the
confinement method~\cite{Tyka2006,Cecchini2009,Ovchinnikov2013} take an alternative strategy for computing difference
free energies that does not require knowing a pathway from $A$ to $B$.

Here, we adopt the confinement method of Tyka et al~\cite{Tyka2006} and Cecchini et al.~\cite{Cecchini2009}, which is
based on the thermodynamic cycle shown in Figure~\ref{fig:method}. We start with an ensemble of related conformations,
$A$. We then restrict $A$ to a much ``tighter'' ensemble $A^\ast$ by applying position restraints in a series of MD
simulations. Similarly, a second ensemble $B$ is restricted to $B^\ast$. We then compute the free energy between
$A^\ast$ and $B^\ast$ using either normal mode analysis \cite{Brooks1983,Case1994} or the quasiharmonic 
method \cite{Karplus1981,Levy1984}. This step takes into account
both the remaining conformational entropy in each ensemble and the remaining enthalpy difference. In this way, we can
calculate the free energy difference between the two end states without needing to define a physical path or reaction
coordinate connecting the two states. The confinement method shares some similarities with the `confine-and-release'
method for computing ligand binding affinities~\cite{Mobley2007,Mobley2012,Mobley2006}.

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{method.pdf}
    \end{center}
    \caption{Graphical representation of the thermodynamic cycle involving confinement method.}
\label{fig:method}
\end{figure}

Previously, the confinement method has been validated on small model peptides~\cite{Tyka2006,Cecchini2009}. Here, we
show that the confinement method can also be applied successfully to larger systems, such as the series of chameleon
proteins designed by Orban and co-workers~\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012}, which can switch between two
completely different folds with only slight changes in sequence. We also show that our computed difference free energies
are useful in evaluating the quality of CASP target protein predictions.  This could be of significant value for
ultimately improving the energetics in protein-structure models, not just the structures. And, finally, we show that we
can approximately decompose difference free energies into individual amino acid components. This offers the opportunity
for diagnosing the structural basis for difference free energies, which can be useful for interpreting biological
mechanisms.

\section*{Results and Discussion}

\subsection*{The confinement method succeeds at some basic consistency checks}

@@

Second, we looked at 6 target proteins from the CASP9 experiment~\cite{Moult2011}.  For each target, we examined up to 5 submitted
models.  We computed the difference free energy between the true native and the best model.
Figure~\ref{fig:summary_casp} and Table~\ref{table:casp_control} show that, in 5 out of 6 cases, the confinement method
assigns a lower free energy to the experimentally determined structure than to any of the decoys.  Other well-known
discriminators can also successfully tell native structures from computer generated models~\cite{Sheffler2009,Zhou2002} 
; it is just an independent useful validation that the confinement calculations make
sense.

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally determined structure than to
    computer-generated predictions. For each target, we examined as many as five predictions submitted by CASP
    participants. We report the free energy difference between the most favorable decoy and the experimentally
    determined structure. Positive $\Delta\Delta G (=G_{best~decoy} - G_{native})$ values indicate that the experimental structure is predicted to be
    more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G$ (kcal/mol) & GDT-TS \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$    &  $44$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$    &  $96$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$    &  $70$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$     &  $94$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$    &  $94$ \\ \hline
     T0569       &    2KYW        &          $20.01 \pm 0.69$    &  $78$ \\ \hline
\end{tabular}
\end{center}
\end{table}

\subsection*{The confinement method correctly predicts the structures of chameleon sequences}

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012}. These are instances in which two highly
similar sequences fold into remarkably different structures. Orban and co-workers have designed a protein-G-like
sequence of 56-residues that is marginally stable in one of two possible folds. By mutating key residues in this
sequence they are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta +
\alpha$ structure as the $\beta$ conformation, and we refer to the $3\alpha$ structure as the $\alpha$ conformation. One
pair of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair
(GA95/GB95) is 95 percent identical, differing in three positions. Accurately predicting the structural preferences of
these structures presents a serious challenge for computational
methods \cite{Allison2011}.

\begin{figure}
    \begin{center}
    \includegraphics[width=5.0in]{orban.pdf}
    \end{center}
    \caption{The confinement method correctly predicts the structural preferences of four chameleon sequences. The top
        part of the figure represent four sequences used in this study along with the protein data bank identifier. The
        experimentally observed fold, computationally predicted fold, and difference free energy between the two folds
        is reported for each sequence.}
\label{fig:orban}
\end{figure}

@@

\subsection*{Per-residue free energy calculations can identify the mechanistic detail behind conformational preferences}

To better understand the mechanism behind these chameleon proteins, we decomposed the calculated free energy into
per-residue contributions in an approximate way~\cite{Tyka2006}. We decompose each confinement step ($\Delta
G_{A,A^\ast}$, $\Delta G_{B,B^\ast}$) into per-residue contributions. We can also decompose the remaining enthalpy in
the confined ensemble into per-residue contributions. However, the method is approximate because we do not include the
residual conformational entropy from the normal mode or quasiharmonic analysis steps. Computing per-residue
contributions helps us to identify important residues that stabilize a particular conformation. The per residue free
energy, $\Delta \Delta G (\beta - \alpha)$ is shown in Figure~\ref{fig:perresidue_orban}.

\begin{figure}
    \begin{center}
        \includegraphics[width=4.5in]{delta_g.pdf}
    \end{center}
    \caption{Per-residue difference free energies for the $\alpha$ and $\beta$ conformations of the GA95 and GB95
        sequences. Residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure,
        and white residues have no preference.}
\label{fig:perresidue_orban}
\end{figure}

Although the overall free energy difference between the two structures is small (within 3--5 kcal/mol), individual
residues can show marked preferences for being in either $\alpha$ or $\beta$ conformation. Such differences can be
understood by looking in detail at the local environment for those residues. For example, the region around residue 7
forms a random coil in the $\alpha$ structure, whereas it is forming a beta sheet structure in the $\beta$ structure.
These residues strongly favour the locally well packed and hydrogen bonded environment found in the $\beta$ sheet.
Overall, favoring $\alpha$ or $\beta$ structure is a delicate balance, where the relative global free energy difference
is small and the contributions of different per residue tendencies balance out. It is therefore very likely that by
changing key residue the global fold preferences can be changed.

@@

\subsection*{The confinement method can correctly rank-order structure predictions}

First, we examined the ability of the confinement method to rank different models for a target that have been generated
by a single group using the same methodology for all predictions. We examined two targets: T0559 and T0560 (see Table~\ref{table:casp_testcase}).

\begin{table}
\begin{center}
\caption{The six different CASP targets that are used in this study. The corresponding PDB Identifier and the 
         organism type are also listed.}
\label{table:casp_testcase}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier &  organism type  \\ \hline
     T0531       &    2KJX        &  extracellular domain of the jumping  \\
                 &                &  translocation breakpoint protein  \\ \hline
     T0538       &    2L09        &  protein asr4154 from Nostoc sp. PCC7120     \\ \hline
     T0540       &    3MX7        &  human Fas apoptotic inhibitory molecule     \\ \hline
     T0559       &    2L01        &  protein BVU3908 from Bacteroides vulgatus   \\ \hline
     T0560       &    2L02        &  protein BT2368 from Bacteroides  \\ 
                 &                &  thetaiotaomicron         \\ \hline
     T0569       &    2KYW        &  domain of adhesion exoprotein from \\ 
                 &                &   Pediococcus pentosaceus \\ \hline
\end{tabular}
\end{center}
\end{table}

The first test case is target T0559. The best predictor group for this 69 amino acid target was ``BAKER-ROSETTASERVER''.
To save computer time, we excluded two models that were very similar to other models that we did include. For this
target, the difference free energy computed by the confinement method can be used to accurately rank-order all of the
models and the native experimental structure (Figure~\ref{fig:T0559}).

\begin{figure}
    \begin{center}
        \includegraphics[width=3.8 in]{T0559.pdf}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and three predictions produced by a single
        method for target T0559.}
\label{fig:T0559}
\end{figure}

We performed a similar calculation for target T0560 with two models from the group ``Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
sate and our calculated free energy based ranking matches well with GDT-TS (Figures S2 in the supporting information).

@@

\subsection*{The per-residue free energy is sensitive to small changes in protein conformation:}

In a previous section, we discussed how the per-residue free energy can reveal mechanistic
detail behind the conformational preference of a chameleon sequence. In this section, our aim is to apply the same
method and try to understand if the per-residue free energy can help us identify residues which stabilize or destabilize
a particular region of a protein. In this case, the two conformations have similar folds with only small changes in
localized regions. We chose CASP target T0569 and compared the experimental NMR structure with the best predicted model
(GDT-TS=78; predicted by the ``Mufold'' group). The confinement method predicts that the experimental structure is more stable
by 20 kcal/mol (Figure S2).

The confinement method clearly identifies two hydrophobic residues V59 and I61, which destabilize the predicted model
with respect to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The sidechains of
these hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but oriented
towards the exterior of the protein and are solvent exposed in the model. These residues are part of a beta-sheet
in the experimental structure, but because of their sidechain orientation the corresponding beta-sheet become disordered in the predicted model 
(Figure~\ref{fig:T0569_per_residue} and Figure S4). There is also a large
difference around K76, which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure.
This suggests that salt-bridge interactions are too favorable for the combination of force field and implicit solvent
model we use, which has been a problem noted in the past \cite{Roe2007}.

@@

\subsection*{The confinement method occasionally produces incorrect results:}

Despite success for most of the studied systems, there are few failures, specially when the GDT-TS scores of the
compared structures are very close. One example is Target T0538, where we compared the experimental structure with three
models (Model 1: ``PconsR''---GDT-TS=96; Model 2: ``Shell''---GDT-TS=90; Model 3: ``FOLDIT''---GDT-TS=86). Contrary to
our expectation, the confinement method predicts that computer generated Model 1 is more stable than the crystal structure (Figure S5).
Per-residue free energy calculations (not shown) show that despite only small variations at the backbone level, the
sidechains are oriented in very different ways (Figure~\ref{fig:T0538compare}), giving rise to large differences in the stabilization of
certain residues. In particular, some of the differences arise from different salt bridge patterns and certain flexible
polar residues exposed to the surface. This unexpected result shows that the confinement method is very sensitive to
local interactions (including sidechain reorientation) and may indicate issues with the forcefield and implicit solvent
models used in the calculation.

\begin{figure}
\begin{center}
\includegraphics[width=4.9 in,height=4.0 in]{Target_538_compare.pdf}
\end{center}
\caption{The per residue free energy calculation found some differences of the sidechain orientation in computer generated model 1 and 
experimental structure of CASP target T0538. 
(A) Two salt bridges between residues E35-R32 and E28-K24 exist in the native structure. These are
absent in the computer generated model structure. Instead
(B) a new salt bridge is formed between residues R32-E28 in model. (C) Another salt bridge between
residues  R26-E50 in the experimental structure. (D) In comparison to the native 
structure, orientation of F51 differs in the model.}
\label{fig:T0538compare}
\end{figure}

\subsection*{Can the confinement method perform better quality assessment in protein structure prediction?}

A part of the CASP experiment is dedicated to the quality assessment (QA) of predicted models \cite{Kryshtafovych2011}.
Here, predictors were asked to score each model on a scale (known as qmode) from 0 to 1, with higher values
corresponding to better models \cite{Kryshtafovych2011}. Most of the groups use consensus strategies in such experiments.
Here, we chose two computer generated models from CASP taget T0538, where the top performing group "MUFOLD-WQA" \cite{Wang2011} failed.
One of the model was from "PconsR" (GDT-TS = 83, qmode 1 = 0.5865) and another from "MULTICOM-NOVEL" ( GDT-TS = 83, qmode 1 = 0.5865). We 
choose these two models as the model by PconsR was the most accurate
predicted model for this target, whereas the other model was predicted best by QA test of MUFOLD-WQA. The calculated
free energy using confinement method indicate that the model by PconsR is more stable by $3.9 Kcal/Mol$ which support
the GDT\_TS trend.  Although consensus prediction strategies do well specially when dealing with large numbers of structures, they 
can miss a good answer that is only predicted by a few methods. At least in this case, the confinement
method is able to capture this structure that was otherwise missed.

\section*{Conclusion}

@@

\item  In the first step, a minimization of A and B conformations are performed. These minimized conformations
    ($A^{\ast}$ and $B^{\ast}$) are the reference conformation of that state.

\item The free energy of confining the ensemble (A or B) to a tighter ensemble ($A^\ast$ or $B^\ast$) is calculated.
    This is done by gradually confining each atom to its position in the reference conformation ($A^{\ast}$ or
    $B^{\ast}$) using a series of progressively stronger position restraints. This is done by running 21 molecular
    dynamics simulation (each 20 ns long) for each leg of the thermodynamic cycle, where the harmonic restraint force
    constant was scaled from 0.00005 Kcal/Mol (mostly free) to 81.92 kcal/mol (tightly restrained). The free energy for
    this step is estimated using a thermodynamic integration approach developed by Tyka et. al. \cite{Tyka2006}. The
    confinement free energies calculated this step are denoted as $\Delta G_{A,A^\ast}$ and $\Delta G_{B,B^\ast}$ in
    Figure~\ref{fig:method}.
